---
title: "R"
output: html_document
date: "2025-06-05"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE, warning = FALSE,fig.width = 10,fig.height = 6)
```

#Packages
```{r libraries}
rm(list = ls())
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(rstudioapi)
library(knitr)
library(kableExtra)

display.brewer.all()

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```


```{r files}
df <- read.csv("titanic.csv")
names(df) <- tolower(names(df))
```


#1

Generate a series of bar charts to describe the gender, ticket class and survival of the passengers onboard
```{r 1}

############## GENDER ####################

gender_colors <- brewer.pal(4, "Set1")[3:4]

#option 1
ggplot(df, aes(x = sex, fill = sex)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..),
            vjust = -0.5,
            color = gender_colors) +
  scale_fill_manual(values = gender_colors) +
  labs(title = "Gender Distribution of the Passengers",
       x = "Gender",
       y = "Count") +
  theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black")
    )

#simpler
ggplot(df, aes(x = sex, fill = sex)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..),
            vjust = -0.5,
            color = gender_colors) +
  scale_fill_manual(values = brewer.pal(4, "Set1")[3:4]) +
  labs(title = "Gender Distribution of the Passengers") +
  theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line.x = element_blank(),
      axis.title.x = element_blank()
    )



############## TICKET CLASS ####################

ticket_colors <- brewer.pal(10, "Spectral")[8:10]

ggplot(df, aes(x = factor(pclass, 
                          levels = c(1, 2, 3), 
                          labels = c("1st", "2nd", "3rd")), 
       fill = factor(pclass))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..),
            vjust = -0.5,
            color = ticket_colors) +
  scale_fill_manual(values = ticket_colors) +
  labs(title = "Ticket Class Distribution of the Passengers",
       x = "Ticket Class",
       y = "Count") +
  theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black")
    )

#simpler
ggplot(df, aes(x = factor(pclass, 
                          levels = c(1, 2, 3), 
                          labels = c("1st", "2nd", "3rd")), 
       fill = factor(pclass))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..),
            vjust = -0.5,
            color = ticket_colors) +
  scale_fill_manual(values = ticket_colors) +
  labs(title = "Ticket Class Distribution of the Passengers") +
  theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line.x = element_blank(),
      axis.title.x = element_blank()
    )


############## SURVIVAL ####################

survival_colors <- brewer.pal(3, "Set1")[1:2]

ggplot(df, aes(x = factor(survived, 
                          levels = c(0, 1),
                          labels = c("Death", "Survived")),
               fill = factor(survived))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..),
            vjust = -0.5,
            color = survival_colors) +
  scale_fill_manual(values = survival_colors) +
  labs(title = "Survival Stutus Distribution of the Passengers",
       x = "Survival Status",
       y = "Count") +
  theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black")
    )

#simpler
ggplot(df, aes(x = factor(survived, 
                          levels = c(0, 1),
                          labels = c("Death", "Survived")),
               fill = factor(survived))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..),
            vjust = -0.5,
            color = survival_colors) +
  scale_fill_manual(values = survival_colors) +
  labs(title = "Survival Status Distribution of the Passengers") +
  theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line.x = element_blank(),
      axis.title.x = element_blank()
    )

```


#2
Generate a histogram for the passengers’ age. 
Furthermore, describe the passengers’ age using the following two boxplots: age per ticket class and age based on survival. 
```{r}
############## HISTOGRAM ####################

ggplot(df, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = brewer.pal(4, "YlOrRd")[3], color = "white") +
  labs(title = "Age Distribution of Passengers",
       x = "Age",
       y = "Count") +
  theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black")
    )


############## BOXPLOT ticket class ####################

ggplot(df, aes(x = factor(pclass, 
                          levels = c(1, 2, 3), 
                          labels = c("1st", "2nd", "3rd")),
               y = age,
       fill = factor(pclass))) +
  geom_boxplot() +
  scale_fill_manual(values = ticket_colors) +
  labs(title = "Passenger Age by Ticket Class",
       x = "Ticket Class",
       y = "Age") +
  theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black")
    )

############## BOXPLOT survival ####################

ggplot(df, aes(x = factor(survived, 
                          levels = c(0, 1),
                          labels = c("Death", "Survived")),
               y = age,
               fill = factor(survived))) +
  geom_boxplot() +
  scale_fill_manual(values = survival_colors) +
  labs(title = "Passanger Age by Survival Status",
       x = "Survival Status",
       y = "Age") +
  theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black")
    )

```


#3
Generate a histogram for the travel fare 
and a table showing the number of people who did not pay 
– you may want to check on Google why a handful of people was on board for free! ???

```{r}
############## HISTOGRAM ####################

ggplot(df, aes(x = fare)) +
  geom_histogram(binwidth = 10, fill = brewer.pal(4, "BuGn")[3], color = "white") +
  labs(title = "Passengers Travelers Fare Distribution",
       x = "Fare Price",
       y = "Count") +
  theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black")
    )

############## TABLE ####################

fare_zero <- df %>%
  filter(fare == 0) %>%
  group_by(pclass, sex, age, survived) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(
    pclass = factor(pclass, levels = c(1, 2, 3), labels = c("1st", "2nd", "3rd")),
    survived = factor(survived, levels = c(0, 1), labels = c("Died", "Survived"))
  )


fare_zero_total <- fare_zero %>%
  summarise(pclass = "Total", sex = "", age = NA, survived = "", Count = sum(Count))
fare_zero_combined <- bind_rows(fare_zero, fare_zero_total)


fare_zero_combined %>%
  kable(col.names = c("Ticket Class", "Sex", "Age", "Survival", "Number of Passengers"),
        caption = "Passengers Who Paid £0 Fare, by Class, Sex, Age, and Survival (with Total)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>% 
  fare_zero_table <- df %>%
  filter(fare == 0) %>%
  group_by(pclass, sex, age, survived) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(
    pclass = factor(pclass, levels = c(1, 2, 3), labels = c("1st", "2nd", "3rd")),
    survived = factor(survived, levels = c(0, 1), labels = c("Died", "Survived"))
  )

# Step 2: Add total row
fare_zero_total <- fare_zero_table %>%
  summarise(
    pclass = "Total",
    sex = "",
    age = NA,
    survived = "",
    Count = sum(Count)
  )

# Step 3: Bind rows together
fare_zero_combined <- bind_rows(fare_zero_table, fare_zero_total)

# Step 4: Display with kable
fare_zero_combined %>%
  kable(col.names = c("Ticket Class", "Sex", "Age", "Survival", "Number of Passengers"),
        caption = "Passengers Who Paid £0 Fare, by Class, Sex, Age, and Survival (with Total)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>% 
  row_spec(0, bold = TRUE, background = brewer.pal(4, "BuGn")[3]) %>% 
  row_spec(nrow(fare_zero_combined), bold = TRUE, background = "#f2f2f2")  


```


#4
A chart of your choice to describe the family size per ticket class

```{r}
############## violin ####################

# parch -> parents/children
# sibsp -> siblings/spouse
# +1 because of the passenger

df <- df %>% 
  mutate(family_size = sibsp + parch + 1)



ggplot(df, aes(x = factor(pclass, 
                          levels = c(1, 2, 3), 
                          labels = c("1st", "2nd", "3rd")),
               y = family_size,
       fill = factor(pclass))) +
  geom_violin(trim = FALSE) +
  scale_fill_brewer(palette = "Pastel2") +
  labs(
    title = "Family Size per Ticket Class",
    x = "Ticket Class",
    y = "Family Size"
  ) +
  theme(
      legend.position = "none",
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_line(color = "black")
    )


```

#5
A series of stacked bar charts to show the how survival differs for different gender and ticket class



